Transient Corotating Clumps Around Adolescent Low-Mass Stars From Four Years of TESS

### 2023-09-05 paper/ms.tex
update ms

% 250 word max!
\begin{abstract}
  We explore a set of pre-main-sequence stars
  with sub-solar masses, rotation periods of 0.2--2 days, optical variability amplitudes
  of a few percent, and sharp recurring light curve features that last for
  hundreds of cycles.  The sharp features are probably caused by
  eclipsing clumps of dust or gas orbiting at the corotation radius.
  We refer to this class of star as ``complex periodic variables'' (CPVs).
  To improve our understanding of these
  enigmatic objects, we conducted a survey for CPVs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CPVs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CPVs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and for the detail with which the
  evolution in each dip's amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CPVs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features could
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CPVs and of B stars
  that may have multipolar magnetic fields.
\end{abstract}

@@

The sharp features of CPV light curves can have durations as short as
5\% of the rotation period ($P_{\rm rot}$), which is too short to be
caused by starspots rotating into and out of view.  Starspots produce
flux variations with characteristic timescales of $P_{\rm rot}$ and
$0.5\,P_{\rm rot}$.  With finely-tuned viewing geometries, starspots
can produce dip durations as short as $\approx$$0.2\,P_{\rm rot}$, but
in such cases, limb darkening causes the dip amplitudes to be smaller
than the observed amplitudes of $\sim$1\% (see
\citealt{2017AJ....153..152S}, Figures 37-41).  Thus, a
``starspot-only'' scenario can be ruled out for many CPVs
\citep{2017AJ....153..152S,2019ApJ...876..127Z,2021MNRAS.500.1366K}.
Given that many CPVs cannot be explained by starspots alone, and
working under the assumption that all CPVs share the same basic
physical scenario, we discard the ``starspot-only'' scenario.
Instead, the correct explanation probably involves material above the
stellar surface that is spatially confined with sharp boundaries
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.

Figure~\ref{fig:f1} illustrates two proposed models for the extrinsic
material.  The first scenario invokes opaque dust ``clumps'' that
orbit near the Keplerian corotation radius [$R_{\rm c} =
(GM/\Omega^2)^{1/3}$, where $\Omega = 2\pi/P_{\rm rot}$] and
periodically transit the star
\citep{2017AJ....153..152S,2023MNRAS.518.4734S}.  The second scenario
invokes ``prominences'', long-lived condensations of cool, dense,
marginally-ionized gas that are embedded within the hotter corona and
that corotate with the star
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
These hypothetical prominences are analogous to quiescent prominences
and filaments seen in the solar corona \citep[see
e.g.][]{2015ASSL..415.....V}, though rather than existing at a
fraction of the stellar radius as in the solar case, they would exist
at distances of a few stellar radii.  A final possibility is that an
optically-thick ring obscures a narrow band of the stellar
photosphere \citep{2019ApJ...876..127Z}.  Hot spots passing behind
such a ring could produce sudden dips.  We do not favor this scenario
(see Appendix~\ref{app:ring}).

The dust clump and gas prominence hypotheses are geometrically
similar; one could imagine dust or gas accumulating in
either of the geometries shown in Figure~\ref{fig:f1}.  The main
distinguishing factor between the two models is the composition of the
occulting material.   ``Dust clumps'' invoke opacity from dust, which
would need to be collisionally charged \citep{2023MNRAS.518.4734S}.
``Gas prominences'' invoke opacity from partially-ionized gas, likely
bound-free transitions in hydrogen or perhaps a molecular opacity.
Unambiguous evidence in support of either scenario has yet to be
acquired.  Such evidence might include a spectroscopic detection of
silicate 10\,$\mu$m dust absorption during a dip, or perhaps detection
of transient Balmer-line excesses as a function of cycle phase,
similar to observations made in systems such as AB~Dor
\citep[see][]{1999ASPC..158..146C} or PTFO 8-8695
\citep{2016ApJ...830...15J}.

In both models, the corotation radius is the location at which matter
concentrates.  The empirical basis for this is that
the sharp CPV features are superposed over smooth, quasi-sinusoidal
starspot profiles.  The theoretical importance of the corotation
radius has been discussed in studies of magnetic rotators
\citep[e.g.][]{1973ApJ...184..271L,1985Ap&SS.116..285N,2005ApJ...634.1214L}.
%
%including neutron stars \citep[e.g.]{1973ApJ...184..271L}, magnetic
%massive stars \citep[e.g.][]{1985Ap&SS.116..285N}, and accreting T
%Tauri stars \citep[e.g.][]{2005ApJ...634.1214L}.  
%
In regions where the magnetic field dominates the flow (i.e.~$B^2/8\pi
> \rho v^2 /2$), matter is dragged along with the field lines.  The
tendency for charged gas or dust to become trapped at corotation is
then tied to how of the four relevant forces -- gravity, Lorentz,
inertial Coriolis, and inertial centrifugal -- the Lorentz and
Coriolis only act perpendicular to field lines, while gravity and the
centrifugal force are in balance at $R_{\rm c}$ \citep[e.g.][their
Section~2]{2005MNRAS.357..251T}.  Another way to phrase this statement
is that there is a local minimum in the effective potential of the
corotating frame; given a flow from either the star or from a
tenuous accretion disk, this local potential minimum enables material
to build up \citep{2005MNRAS.357..251T}.

Although theoretical heritage for understanding rapid magnetic
rotators exists, CPVs have remained mysterious because they have been both
hard to discover and hard to characterize.   They have been hard to discover
because they are rare: CPVs comprise $\approx$1\% of the youngest
$\approx$1\% of M dwarfs \citep{2018AJ....155..196R}.  Out of the
millions of stars monitored by K2 and TESS, about 50 CPVs have been
reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2021AJ....161...60S,2022AJ....163..144G,2023ApJ...945..114P}.
They have been hard to characterize because many of the known CPVs are
faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, with optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  At such magnitudes, high-resolution
time-series spectroscopy is out of reach with current facilities,
despite the potential utility of such observations.

@@

Some comments on nomenclature are needed.  What we are calling
``complex periodic variables'' \citep{2023MNRAS.518.2921K} have also
been called ``complex rotators''
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P},
``transient flux dips'', ``persistent flux dips'', and ``scallop
shells'' \citep{2017AJ....153..152S}.  The CPVs should not be conflated
with ``dippers'', which are classical T Tauri stars with infrared
excesses, and which show large-amplitude variability linked to
obscuring inner disk structures and accretion hot spots
\citep{2014AJ....147...82C,2021ApJ...908...16R}.  The phenomenology
and stellar properties of CPVs and dippers are quite different (though
see Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}).  The
defining phenomenological features of the CPVs are that their
light curves are {\it complex}, relative to quasi-sinusoidal
starspots, and the complex features are {\it periodic}, over
timescales of order one hundred cycles.  While rotation likely
does play a central role in explaining their physical behavior, the
acronym for ``complex rotator'' is already widely used in the astrophysical
literature for cosmic rays.  Given these
considerations, we refer to the stars as complex periodic variables
(CPVs); our preferred explanation for their behavior is that transient
clumps of gas or dust orbit at the corotation radius.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.46\textwidth]{f7a.pdf}
			\includegraphics[width=0.46\textwidth]{f7b.pdf}
		}
		\vspace{-0.2cm}

		\subfloat{
			\includegraphics[width=0.46\textwidth]{f7d.pdf}
			\includegraphics[width=0.46\textwidth]{f7c.pdf}
		}
	\end{center}
	\vspace{-0.4cm}
	\caption{
    {\bf River plots of the LP 12-502 light curve}, showing (clockwise
    from top-left) Sectors 18-19, 25-26, 53, and 58-59.  A
    two-harmonic sinusoid has been subtracted to highlight the sharp
    dips.  
    A fixed period and phase are adopted for all sectors; the dips
    across all observations are bounded by $\phi \in [-0.35,0.35]$.
    In Sectors 25-26 (cycles 248-315), periods are visible at
    the fundamental period of 18.5611\,hr, as well as at faster
    ($\phi$$\approx$0-0.07) and slower ($\phi$$\approx$0.25-0.27)
    relative periods based on the presence of blue dips with distinct
    slopes.  Multiple simultaneous periods are also visible in Sector
    53 (cycles 1234-1263) and Sectors 58-59 (cycles 1411-1479).  White
    chunks denote missing data.
    The state changes noted in red in Figure~\ref{fig:lplc} occur in
    cycles 261, 309, and 1241.
	}
	\label{fig:lpriver0}
\end{figure*}

>>>>

% 250 word max!
\begin{abstract}
  We explore a set of pre-main-sequence stars
  with sub-solar masses, rotation periods of 0.2--2 days, optical variability amplitudes
  of a few percent, and sharp light curve features that repeat for
  hundreds of rotation cycles.  The sharp features are probably caused by
  eclipsing clumps of dust or gas orbiting at the corotation radius.
  We refer to this class of star as ``complex periodic variables'' (CPVs).
  To improve our understanding of these
  enigmatic objects, we conducted a survey for CPVs using TESS
  short-cadence data collected between 2018~July and 2022~September.
  Our search of \nstarssearched\ K and M dwarfs with $T$$<$16 and
  $d$$<$150\,pc yielded \ngoods\ high-quality CPVs, most of which are
  new discoveries.  Our sample includes the brightest
  ($T$$\approx$9.5), closest ($d$$\approx$20\,pc), and oldest
  ($\approx$200\,Myr) known CPVs.  One of the new discoveries, LP
  12-502, stands out for the complexity of its light curve --- with as
  many as eight flux dips per cycle --- and for the detail with which 
  each dip's evolving amplitude and phase can be tracked over
  1{,}500 cycles.  We argue that corotating gas or dust remains the
  most viable explanation for CPVs. The gas or dust is probably
  entrained by the star's magnetic field, and the sharp features could
  result from a multipolar field topology, a hypothesis supported by
  correspondences between the light curves of CPVs and of B stars
  that may have multipolar magnetic fields.
\end{abstract}

@@

The sharp features of CPV light curves can have durations as short as
5\% of the rotation period ($P_{\rm rot}$), which is too short to be
caused by starspots rotating into and out of view.  Starspots produce
flux variations with characteristic timescales of $P_{\rm rot}$ and
$0.5\,P_{\rm rot}$.  With finely-tuned viewing geometries, starspots
can produce dip durations as short as $\approx$$0.2\,P_{\rm rot}$, but
in such cases, limb darkening causes the dip amplitudes to be smaller
than the observed amplitudes of $\sim$1\% (see
\citealt{2017AJ....153..152S}, Figures 37-41).  Thus, a
``starspot-only'' scenario can be ruled out for many CPVs
\citep{2017AJ....153..152S,2019ApJ...876..127Z,2021MNRAS.500.1366K}.
Given that many CPVs cannot be explained by starspots alone, and
working under the assumption that all CPVs share the same basic
physical scenario, we discard the ``starspot-only'' scenario.
Instead, the correct explanation probably involves spatially concentrated
circumstellar material
\citep[e.g.][]{2017AJ....153..152S,2022AJ....163..144G}.

Figure~\ref{fig:f1} illustrates two proposed models for the extrinsic
material.  The first scenario invokes opaque dust ``clumps'' that
orbit near the Keplerian corotation radius [$R_{\rm c} =
(GM/\Omega^2)^{1/3}$, where $\Omega = 2\pi/P_{\rm rot}$] and
periodically transit the star
\citep[e.g.][]{2017AJ....153..152S,2017MNRAS.471L.145F,2023MNRAS.518.4734S}.  The second scenario
invokes ``prominences'', long-lived condensations of cool, dense,
marginally-ionized gas that are embedded within the hotter corona and
that corotate with the star
\citep{1989MNRAS.238..657C,2019MNRAS.482.2853J,2022MNRAS.514.5465W}.
These hypothetical prominences are analogous to quiescent prominences
and filaments seen in the solar corona \citep[see
e.g.][]{2015ASSL..415.....V}, though rather than existing at a
fraction of the stellar radius as in the solar case, they would exist
at distances of a few stellar radii.  A final possibility is that an
optically-thick ring obscures a narrow band of the stellar
photosphere \citep{2019ApJ...876..127Z}; hot spots passing behind
such a ring could produce sudden dips.  We disfavor this scenario
for reasons described in Appendix~\ref{app:ring}.

The dust clump and gas prominence hypotheses are conceptually
similar; one could imagine dust or gas accumulating in
either of the geometries shown in Figure~\ref{fig:f1}.  The main
distinguishing factor between the two models is the composition of the
occulting material.   ``Dust clumps'' invoke opacity from dust, which
would need to be collisionally charged \citep{2023MNRAS.518.4734S}.
``Gas prominences'' invoke opacity from partially-ionized gas, likely
bound-free transitions in hydrogen or perhaps a molecular opacity.
Unambiguous evidence in support of either scenario has yet to be
acquired.  Such evidence might include a spectroscopic detection of
silicate 10\,$\mu$m dust absorption during a dip, or perhaps detection
of transient Balmer-line excesses as a function of cycle phase,
similar to observations made in systems such as AB~Dor
\citep[see][]{1999ASPC..158..146C} or PTFO 8-8695
\citep{2016ApJ...830...15J}.

In both models, the corotation radius is the location at which matter
concentrates.  The empirical basis for this is that
the sharp CPV features are superposed over smooth, quasi-sinusoidal
starspot profiles.  The theoretical importance of the corotation
radius has been noted in previous studies of magnetic rotators
\citep[e.g.][]{1973ApJ...184..271L,1985Ap&SS.116..285N,1991ApJ...370L..39K,2005ApJ...634.1214L}.
%
%including neutron stars \citep[e.g.]{1973ApJ...184..271L}, magnetic
%massive stars \citep[e.g.][]{1985Ap&SS.116..285N}, and accreting T
%Tauri stars \citep[e.g.][]{2005ApJ...634.1214L}.  
%
In regions where the magnetic field dominates the flow (i.e.~$B^2/8\pi
> \rho v^2 /2$), matter is dragged along with the field lines.  
Within such regions, charged gas or dust can become trapped at corotation because
of the four relevant forces -- gravity, Lorentz, inertial
Coriolis, and inertial centrifugal -- the Lorentz and Coriolis only
act perpendicular to field lines, while gravity and the centrifugal
force are in balance at $R_{\rm c}$ \citep[e.g.][their
Section~2]{2005MNRAS.357..251T}.  Another way to phrase this statement
is that, in the corotating frame, the effective potential experienced
by charged particles tends to have local minima at $R_{\rm c}$; given
a flow from either the star or from a tenuous accretion disk, this
local potential minimum enables material to build up
\citep{2005MNRAS.357..251T}.

While theoretical heritage for understanding magnetic
rotators exists, CPVs have remained mysterious because they have been both
hard to discover and hard to characterize.   They have been hard to discover
because they are rare: CPVs comprise $\approx$1\% of the youngest
$\approx$1\% of M dwarfs \citep{2018AJ....155..196R}.  Out of the
millions of stars monitored by K2 and TESS, about 50 CPVs have been
reported to date
\citep{2016AJ....152..114R,2017AJ....153..152S,2018AJ....155...63S,2019ApJ...876..127Z,2020AJ....160...86B,2021AJ....161...60S,2022AJ....163..144G,2023ApJ...945..114P}.
They have been hard to characterize because many of the known CPVs are
faint; the initial K2 discoveries
\citep{2016AJ....152..114R,2017AJ....153..152S} were M2-M6 dwarfs at
distances $\gtrsim$100\,pc, with optical brightnesses of
$V$$\approx$15.5 to $V$$>$20.  At such magnitudes, high-resolution
time-series spectroscopy is out of reach with current facilities,
despite the potential utility of such observations.

@@

Some comments on nomenclature are needed.  What we are calling
``complex periodic variables'' \citep{2023MNRAS.518.2921K} have also
been called ``complex rotators''
\citep{2019ApJ...876..127Z,2022AJ....163..144G,2023ApJ...945..114P},
``transient flux dips'', ``persistent flux dips'', and ``scallop
shells'' \citep{2017AJ....153..152S}.  The CPVs should not be conflated
with ``dippers'', which are classical T Tauri stars with infrared
excesses, and which show large-amplitude variability linked to
obscuring inner disk structures and accretion hot spots
\citep{2014AJ....147...82C,2021ApJ...908...16R}.  The phenomenology
and stellar properties of CPVs and dippers are quite different (though
see Sections~\ref{subsec:irexcess} and~\ref{subsec:discdippers}).  The
defining phenomenological features of the CPVs are that their
light curves are {\it complex}, relative to quasi-sinusoidal
starspots, and the complex features are {\it periodic}, over
timescales of order one hundred cycles.  While rotation likely
does play a central role in explaining their physical behavior, the
acronym for ``complex rotator'' is already used in the astrophysical
literature for cosmic rays.  Given these
considerations, we refer to the stars as complex periodic variables
(CPVs); our preferred explanation for their behavior is that transient
clumps of gas or dust orbit at their corotation radii.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{f7a.pdf}
			\includegraphics[width=0.48\textwidth]{f7b.pdf}
		}
		\vspace{-0.2cm}

		\subfloat{
			\includegraphics[width=0.48\textwidth]{f7d.pdf}
			\includegraphics[width=0.48\textwidth]{f7c.pdf}
		}
	\end{center}
	\vspace{-0.4cm}
	\caption{
    {\bf River plots of the LP 12-502 light curve}, showing (clockwise
    from top-left) Sectors 18-19, 25-26, 53, and 58-59.  A
    two-harmonic sinusoid has been subtracted to highlight the sharp
    dips.  
    A fixed period and phase are adopted for all sectors; the dips
    across all observations are bounded by $\phi \in [-0.35,0.35]$.
    In Sectors 25-26 (cycles 248-315), periods are visible at
    the fundamental period of 18.5611\,hr, as well as at faster
    ($\phi$$\approx$0-0.07) and slower ($\phi$$\approx$0.25-0.27)
    relative periods based on the presence of blue dips with distinct
    slopes.  Multiple simultaneous periods are also visible in Sector
    53 (cycles 1234-1263) and Sectors 58-59 (cycles 1411-1479).  White
    chunks denote missing data.
    The state changes noted with red markers in Figure~\ref{fig:lplc} occur in
    cycles 261, 309, and 1241.
	}
	\label{fig:lpriver0}
\end{figure*}
